---
title: "Visualizing Census Data for the West Coast"
author: "aaron mamula"
date: "3/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

This short document is meant to provide a minimal reproduceable example of how to create some basic chloropleath maps using Census Data.

This Vignette assumes that you have already used the functions from [svi_censustract_functions.R](https://github.com/aaronmams/svi-tools/blob/master/R/SVI_censustract_functions.R) to create the Census Tract level data frame with the 16 social vulnerability indicators.

## Dependencies

### R packages

The mapping exercises here rely primarily on methods from the 

* ggmap
* maptools

packages.  In order to deal with shapefiles you may also require the use of two other packages:

* rgdal
* rgeos

### Additional data

The mapping exercises here rely on a couple of shapefiles that I downloaded from the [Census Bureau's Tiger Data Mart](https://www.census.gov/geo/maps-data/data/tiger-cart-boundary.html). The two that I will be using here are:

1. "gz_2010_06_140_00_500k" - the cartographic boundaries for California Census Tracts, and
2. "gz_2010_06_060_00_500k" - cartographic boundaries for California Counties.

## Exercise 1: Create a 'heatmap' of poverty by Census Tract

The first step here is to set up the social vulnerability indicator data.  As explained in [this Vignette](https://github.com/aaronmams/svi-tools/blob/master/svi_census_tract.Rmd) I don't really want to wait 30 minutes for the data to compile from API pings so I'm just going to load a version of the data that I stored:

```{r}
svi.df <- readRDS('data/svi_censustract_2015.RDA')
```

```{r warning=FALSE, message=FALSE}
library(rgdal)
library(ggplot2)
library(ggmap)
library(scales)
library(maptools)
library(rgeos)
library(dplyr)

tract <- readOGR(dsn = "shapefiles/gz_2010_06_140_00_500k", 
                 layer = "gz_2010_06_140_00_500k")
tract <- fortify(tract, region="GEO_ID")


ca.poverty <- tbl_df(svi.df) %>%
              select(name,state,county,tract,pov.pct) %>%
              filter(state=='06') %>%
              mutate(id=paste('1400000US',state,county,tract,sep=""))

plotData <- left_join(tract,ca.poverty)

county <- readOGR(dsn = "shapefiles/gz_2010_06_060_00_500k", 
                  layer = "gz_2010_06_060_00_500k")
county <- fortify(county, region="COUNTY")

#now I'm going to use the scales() package to bin the data a little
# so the contrasts in coverage rates are a little more visible
ggplot() +
  geom_polygon(data = plotData, aes(x = long, y = lat, group = group,
                                    fill = pov.pct)) +
  geom_polygon(data = county, aes(x = long, y = lat, group = group),
               fill = NA, color = "black", size = 0.25) +
  coord_map() +
  scale_fill_distiller(palette = "Greens", labels = percent,
                       breaks = pretty_breaks(n = 10)) +
  guides(fill = guide_legend(reverse = TRUE)) + theme_bw()

```

## Exercise 2: heatmap of uninsured rates for CA North Coast

